Selection on early survival does not explain germination rate clines in *Mimulus cardinalis* (Phrymaceae)

### 2022-06-23 ms/ms.Rmd
r1 prerelease

*short title*: Selection on germination in *Mimulus cardinalis*

@@

fit <- read_rds("r/objects/fit.rds")
```

@@

**Conclusions** Southern populations have higher early survival but this is not caused by faster germination. This pattern is consistent with nonadaptive forces driving the phenotypic cline in germination rate, but future work will need to assess whether there is selection at other life stages. The statistical framework should help expand quantitative genetic analyses for other waiting-time traits.

@@

<!--- broad problem: local adaptation --->
Populations within species differ in ecologically important traits that often result from adaptation to different local environments [@turesson_genotypic_1922; @clausen_experimental_1948]. However, it is still rare that we understand the traits and selective agents underlying local adaptation [@wadgymar_identifying_2017]. The classic signature of local adaptation is crossing reaction norms for fitness measured in a reciprocal transplant experiment [@kawecki_conceptual_2004; @johnson_reciprocal_2021], such that populations have  higher relative fitness in their local environment but lower relative fitness in a foreign environment. For example, annual inland ecotypes of the yellow Monkeyflower *Mimulus guttatus* have higher relative fitness in seasonally dry sites compared to perennial coastal ecotypes that are adapted to year-round water availability. The relative fitness of coastal and inland ecotypes is reversed on the coast because of life history and other genetic differences between ecotypes [@lowry_widespread_2010]. Local adaptation is important in evolutionary biology because its prevalence implies that selection varies spatially, populations have heritable variation in fitness, and selection is stronger than migration or drift. Exceptional cases of local adaptation over short temporal or spatial scales (e.g. @wright_indirect_2013; @grant_40_2014; @richardson_microgeographic_2014; @lescak_evolution_2015; @barrett_linking_2019; @divittorio_natural_2020) likely contribute to an availability bias that leads many to perceive that local adaptation is ubiquitous. However, systematic meta-analyses reveal that local adaptation is often weak or nonexistent [@leimu_meta-analysis_2008; @hereford_quantitative_2009; @brady_understanding_2019] and may be getting weaker because of anthropogenic climate change [@bontrager_climate_2020]. Populations may not be locally adapted if differential selection is weak relative to migration or drift or they lack heritable variation in traits under selection. Alternatively, recent anthropogenic climate change may have erased the signature of local adaptation by causing some foreign populations to have higher fitness than local populations [@bontrager_climate_2020]. For example, in the alpine plant *Boechera strica*, simulated climate change (mimicking early snow melt) favors populations from lower elevations and snow addition restores the relative fitness advantage of high-elevation populations [@anderson_climate_2020]. Understanding what causes trait variation among populations within species, whether adaptive or not, will help biologists better predict how populations will respond to environmental change.

<!--- narrow problem: are phenotypic clines good evidence for local adaptation? --->
Phenotypic clines are commonly interpreted as evidence for local adaptation, but nonadaptive processes also generate similar patterns [@endler_geographic_1977]. A phenotypic cline is a correlation between a heritable character and the environment or a proxy for environmental gradients such as latitude or elevation [@huxley_clines_1938]. Clines may be caused by either abrupt or gradual environmental change. For example, heavy metal contamination of soil around mines is an abrupt change from the surrounding pasture soil. There is an abrupt change in the heavy metal tolerance of *Anthoxanthum odoratum* populations across the mine boundary, but other traits under weaker or correlated selection vary gradually over space [@antonovics_evolution_1970]. Other clines track gradual environmental change. Latitudinal clines in the size of *Drosophila subobscura* flies have evolved multiple times and most likely track gradual variation in temperature and/or phenology. But clines are not necessarily adaptive. When there is genetic isolation by distance [@wright_isolation_1943], it is possible for phenotypic differences to track genetic differences throughout the species range [@vasemagi_adaptive_2006]. In practice, nonadaptive clines are difficult to demonstrate because it is hard to reject adaptive explanations. The proportion of cyanogenesis varies clinally in white clover (*Trifolium repens*) but field experiments have not observed spatially varying selection which could explain this cline [@wright_genetic_2021]. Hence, spatial variation in cyanogenesis may be nonadaptive or the design of the experiment may have missed selection occurring during germination and early development. When species expand their range, such as during biological invasions, nonadaptive clines can readily evolve because of multiple introductions and serial founder events [@colautti_contemporary_2015]. Even parallel clines, which are usually considered strong evidence of natural selection, can result from nonadaptive processes when there is epistasis and consistent spatial variation in the strength of genetic drift [@santangelo_modern_2018]. Evolutionary ecological experiments such as reciprocal transplants are needed to provide additional evidence to distinguish between adaptive and nonadaptive hypotheses for clines [@wadgymar_integrating_2017].

There are several experimental methods for measuring the agents and strength of selection to different environments under natural conditions [@wadgymar_identifying_2017], but here we focus on differential selection. One key prediction is that the direction of selection and position of the phenotypic optimum will differ between environments. If genetic variation in a trait *causes* fitness to vary, then genotypic selection analyses [@rausher_measurement_1992] should find that genotypes with high trait expression are favored in one environment but disfavored in another. For example, selection favors higher seed dormancy in southern environments of *Arabidopsis thaliana* and lower dormancy in northern environments, presumably to optimally time seedling establishment with the onset of favorable summer conditions [@postma_early_2016]. Populations can also simultaneously differ as a result of both adaptive and non-adaptive processes. When locally adapted populations differ in traits that do not themselves confer local adaptation, we expect population-level covariation between divergent traits and fitness but do not expect within-population variation to correlate with fitness. Hence, comparing the direction of among-genotype and among-population selection can help distinguish between phenotypic clines that are important for local adaptation and those which may have evolved nonadaptively.

In this study, we used a reciprocal transplant to test whether a cline in *Mimulus cardinalis* germination rate could be locally adaptive in terms of early survival between seedling establishment and flowering. Because seeds of *M. cardinalis* do not require cold or warm stratification to break dormancy, they will germinate whenever conditions are sufficiently warm and moist.  Within this germination niche, there is genetic variation for germination time that covaries with latitude and other traits associated with fast-slow growth continuum [@muir_grow_2017; @sheth_demographic_2018; @nelson_quantitative_2021]. In greenhouse conditions, seeds from the southern end of the range (San Diego County, CA) germinate in 6-7 days; seeds from the northern end of the range (Lane County, OR) germinate in 9-10 days. The adaptive hypothesis is that faster germination increases early survival in the southern end of the range and slower germination increases early survival at the northern end. We would predict either stabilizing genotypic selection for different optimal germination rates or opposing signs of directional genotypic selection. If the cline is maintained by nonadaptive forces then germination rate should be uncorrelated with genotypic fitness (even if it is correlated with population mean fitness) or the direction of selection should be the same in both parts of the range. A limitation of this study is that we cannot measure selection on seedling establishment since that occurred in the greenhouse (see [Materials and Methods]) or later survival and fecundity (these fitness components will be analyzed in a follow-up paper).

Germination and early survival are potentially important for local adaptation but often overlooked compared to later life stages. Failure to measure seedling establishment and early survival can bias estimates of selection on traits that only express at later life stages, such as flower size [@mojica_viability_2010]. In the herb *Arabidopsis thaliana*, quantitative differences in germination timing contributed to local adaptation and are maintained by a balance of selection on seedling establishment, survival, and fecundity, but the optimal timing varies in northern and southern environments [@postma_early_2016; @postma_among-year_2018]. A challenge of measuring genotypic selection on germination rate is that quantitative genetic theory and analysis are built around Gaussian distributions [@lynch_genetics_1998; but see @de_villemereuil_general_2016], but the distribution of times until germination are most likely not Gaussian. Note that time to germination is the inverse of germination rate. The time until germination can be modeled as a waiting time, which are bounded at zero and usually right skewed. Waiting times are distributed exponentially if the probability of germination is constant through time, but this unlikely to be true. Rather, the probability of germination almost certainly increases through time once conditions are conducive. Hence, more complex waiting time distributions should be seen. A second challenge is that waiting time is usually modeled as a continuous trait, but is observed at discrete intervals (e.g. once per day). If the average time to germination is sufficiently long relative to the interval between observations, then a continuous-time model may be sufficient. However, when time to germination is several days, as with *M. cardinalis*, then accurately modeling the trait may need to account for the discrete observation process. Hence, investigating spatially variable selection on germination time is important for expanding our knowledge of local adaptation and phenotypic clines for poorly studied life stages with challenging statistical properties.

@@

*Mimulus cardinalis* ($=$*Erythranthe cardinalis* [@lowry_case_2019]) is a perennial forb native to the Western US (California and Oregon). It is predominantly outcrossing, self-compatible, hummingbird-pollinated. We used five focal populations from throughout the geographic range of *M. cardinalis* (Table \ref{tab:focal_pops}). These five focal populations are a subset of those we used previously to identify a cline in germination and other traits [@muir_grow_2017]. Seeds were collected in the field from mature, undehisced fruits left in open coin envelopes for 2-4 weeks to dry, then stored at room temperature. To control for maternal effects, we grew a large number of field-collected seeds in the greenhouse and generated seed families for this experiment by hand-pollinating individuals using the breeding design described in the next section.

@@

tribble(
  ~Name, ~Latitude, ~Longitude, ~`Elevation (mas)`,
  "Sweetwater River", 32.900, -116.585, 1180,
  "West Fork Mojave River", 34.284, -117.378, 1120,
  "North Fork Middle Tule River", 36.201, -118.759, 926,
  "Little Jamison Creek", 39.743, -120.704, 1603,
  "Rock Creek", 43.374, -122.957, 326
) |>
  knitr::kable(format = "markdown", caption = "\\label{tab:focal_pops} Source populations, including the name of the drainage where the seeds were collected, the latitude, longitude, and elevation in meters above sea level (mas).", align = "lccc", label = NA)
```

@@

To estimate selection among genotypes and populations, we first need to quantify genetic variance in germination rate and early survival. We estimated genetic variance and heritability of germination rate and winter survival using a quantitative genetic breeding design. For each population, we crossed 15 parental individuals using a partial diallel design [@lynch_genetics_1998] with three dams per sire and three sires per dam for a total of 45 full-sib families per population. Individual plants of this self-compatible species were used as both sires (pollen parent) and dams (ovule parent). We did not make crosses between populations. One family from the Little Jamison Creek population did not produce enough seeds, resulting in 44 usable families. In total, we used $5 \times 45 - 1 = 224$ families.

We estimated genetic variances and heritabilities for each of the five populations (Table \ref{tab:focal_pops}). The total phenotypic variance within a population $V_P$ can be partitioned into genetic ($V_G$) and environmental ($V_E$) components:

$$V_P = V_G + V_E$$
We did not have statistical power to estimate nonadditive genetic contributions, therefore we made the simplifying operational assumption that all genetic variance is additive. The broad-sense ($H^2$) heritability is the fraction of phenotypic variance contributed by additive genetic variance:

@@

$$V_G = 4~\sigma_{\mathrm{parent}}$$
Using $\sigma_{\mathrm{parent}}$ rather than $\sigma_{\mathrm{sire}}$ is more powerful because it uses all of the data, but assumes that maternal and paternal effects are negligible. We estimated $\sigma_{\mathrm{parent}}$ and other parameters using a Bayesian mixed effects model. We describe the general approach here and provide detail specific to germination rate and survival below. We fit the model in *Stan* `r str_extract(as.character(fit$version$cmdstan), "^2.[1-2][0-9]")` [@stan_development_team_stan_2021] using **cmdstanr** version `r as.character(fit$version$cmdstanr)` [@gabry_cmdstanr_2021]. *Stan* calculates the posterior distribution of the model using the Hamiltonian Monte Carlo algorithm, which is similar to the more widely used Markov Chain Monte Carlo, but is faster and more efficient at sampling for many applications [@monnahan_faster_2017]. We used weakly informative normally distributed priors for parameters that affect the trait mean (intercepts and coefficients) and half-Cauchy priors for variance parameters. Weakly informative priors are strongly recommended for complex mixed models [@mcelreath_statistical_2016]. We ran `r length(fit$metadata()$id)` parallel chains for `r fit$metadata()$iter_warmup` warmup iterations, `r fit$metadata()$iter_sampling` sampling iterations, and a thinning interval of `r fit$metadata()$thin`. This configuration allowed parameters to converge, which we defined as the convergence diagnostic $\hat{R} < 1.05$ and an effective sample size $>1000$ [@vehtari_rank-normalization_2021]. We inspected posterior predictive plots to assess model adequacy using the `pp_check` function from the *R* package **bayesplot** version `r packageVersion("bayesplot")` [@gabry_visualization_2019]. We used the posterior median for point estimates and calculated uncertainty with the 95\% quantile intervals from the posterior distribution. *Stan* and *R* code are available on GitHub ([https://github.com/cdmuir/mimulus-germination](https://github.com/cdmuir/mimulus-germination)) and will be archived on Zenodo upon publication. Data will be deposited on Dryad upon publication.

@@

We measured germination rate by counting the number of days to germination in 48 randomized blocks split evenly between two time cohorts (North cohort: August 3-4 and South cohort: August 22-24, 2015) in the University of British Columbia greenhouse (Vancouver, BC, Canada). The North and South designations refer to the transplant garden location, as described below. We sowed 3-5 seeds per plug on moist perlite-peat moss potting medium (Sunshine #1, Sungro Horticulture, Agawam, Massachusetts, USA) and recorded the first day on which a plug had at least one germinated seedling. We later thinned each plug to one seedling by selecting the central-most individual, so the first germinated seedling recorded is generally not the same individual transplanted into the field. Hence, individual correlations between germination and survival are inappropriate and we analyze relationships between breeding values or population averages. We censused germination daily for most of the experiment, but census dates were more spread out later in the experiment when only a few plants remained. We accounted for census interval in our statistical model (see below). Gentle misting kept soil moist for the duration of the experiment. Most plugs (`r length(which(data$Germinate == "Y"))` of `r nrow(data)`, `r round(100 * length(which(data$Germinate == "Y")) / nrow(data), 1)`%) had at least one germinant by the end of the experiment (North cohort: August 27, 2015; South cohort: September 20, 2015); we treated plants that did not germinate as missing data. `r nRemove` (`r round(100 * nRemove / length(which(data$Germinate == "Y")), 2)`%) plants emerged before day 5. We removed data on these individuals because they were likely not *M. cardinalis*, but rather contamination from another species' seed in the potting medium. Species can be difficult to tell apart as the cotyledons first emerge, but in plugs where a single germinant emerged before day 5 it was always a contaminant species. It therefore seems most likely that all germinants before day 5 were contamination, but observations during thinning showed that contamination was rare and it unlikely to have any substantial impact on our estimate of germination timing.

## Estimating genetic variance in germination rate
The full germination-rate model included a fixed effect of population, fixed effect of cohort, a random effect of greenhouse block, and a random effect of parent as either dam or sire to estimate additive genetic variance. Most quantitative traits are modeled with a Gaussian (i.e. normal) distribution, but this is inappropriate for germination which has a lower bound at 0 and is censused at discrete time points. A Gaussian distribution might be an adequate approximation if the time to germination was longer and more spread out so that the effects of a zero lower bound and discrete sampling interval were inconsequential. However, given the rapid germination rate and small spread in timing for *M. cardinalis*, we used a different approach. Instead, we modeled genetic variation in a continuously distributed latent rate parameter that is measured as a discrete number of days to germination. Time to germination can be modeled similar to other waiting time or survival processes. There are several common waiting-time distributions with a lower bound at zero, but we used the log-normal distribution because unlike some other common distributions (e.g. exponential), the mean and variance of log-normal distribution are separate parameters. Hence, we can model the variance components separate from differences in the mean germination rate. We discretized the distribution using the definite integral of the probability density to calculate a probability mass. Let the probability density with parameters $\theta$ be $f(x|\theta)$. If a seed was censused to have germinated on day $i$, it could have germinated at any time between the previous census on day $t_{i-1}$. The associated probability mass $g(x|\theta)$ is:

$$g(x=t_i|\theta) = \int_{t_{i-1}}^{t_i} f(x|\theta) dx$$

Preliminary analyses revealed a germination threshold of four days since we did not observe any germination before the fifth day after sowing. The model did not fit as well without this threshold because models without a threshold predict a significant amount of germination between days 0 and 5 (results not shown). The variances and heritability are reported on the log-tranformed scale.

@@

We compared the climate during the experiment (September 2015 to May 2016) to a standardized climate normal. Following [@sheth_demographic_2018], we downloaded climate variables for 1961-1990 and 2015-16 derived from ClimateNA version 7.10 [@wang_locally_2016]. We compared the seasonal temperature average (autumn: September-November; winter: December-February; spring: March-May) for the experiment with the 1961-1990 normal for each population.

@@

## Southern populations germinate faster and have a higher probability of winter survival

@@

Fig. S1-S2 and Tables S1-S4 may be found online in the Supporting Information section.

@@

>>>>

*Short title*: Selection on germination in *Mimulus cardinalis*

@@

fit <- read_rds("r/objects/fit.rds")

# estimate and CI of change in log-mean days to germination
germ_north_diff <- fit$draws("bCohortNorth_germ") |>
  posterior::as_draws_df() |>
  ggdist::point_interval(.point = median, .interval = tidybayes::qi) |>
  dplyr::mutate(dplyr::across(where(is.numeric), round, digits = 3)) |>
  dplyr::mutate(median_qi = glue::glue("${bCohortNorth_germ}~({.lower}, {.upper})$")) |>
  dplyr::pull(median_qi)

@@

**Conclusions** Southern populations have higher early survival but this is not caused by faster germination. This pattern is consistent with nonadaptive forces driving the phenotypic cline in germination rate, but future work will need to assess whether there is selection at other life stages. This statistical framework should help expand quantitative genetic analyses for other waiting-time traits.

@@

<!--- broad problem: local adaptation --->
Populations within species differ in ecologically important traits that often result from adaptation to different local environments [@turesson_genotypic_1922; @clausen_experimental_1948]. However, it is still rare that we understand the traits and selective agents underlying local adaptation [@wadgymar_identifying_2017]. The classic signature of local adaptation is crossing reaction norms for fitness measured in a reciprocal transplant experiment [@kawecki_conceptual_2004; @johnson_reciprocal_2021], such that populations have  higher relative fitness in their local environment but lower relative fitness in a foreign environment. For example, annual inland ecotypes of the yellow Monkeyflower *Mimulus guttatus* have higher relative fitness in seasonally dry sites compared to perennial coastal ecotypes that are adapted to year-round water availability. The relative fitness of coastal and inland ecotypes is reversed on the coast because of life history and other genetic differences between ecotypes [@lowry_widespread_2010]. The prevalence of local adaptation implies that selection varies spatially, populations have heritable variation in fitness, and selection is stronger than migration or drift. Exceptional cases of local adaptation over short temporal or spatial scales [e.g. @wright_indirect_2013; @grant_40_2014; @richardson_microgeographic_2014; @lescak_evolution_2015; @barrett_linking_2019; @divittorio_natural_2020] likely contribute to an availability bias that leads many to perceive that local adaptation is ubiquitous. However, systematic meta-analyses reveal that local adaptation is often weak or nonexistent [@leimu_meta-analysis_2008; @hereford_quantitative_2009; @brady_understanding_2019] and may be getting weaker because of anthropogenic climate change [@bontrager_climate_2020]. Populations may not be locally adapted if differential selection is weak relative to migration or drift or they lack heritable variation in traits under selection. Alternatively, recent anthropogenic climate change may have erased the signature of local adaptation by causing some foreign populations to have higher fitness than local populations [@bontrager_climate_2020]. For example, in the alpine plant *Boechera strica*, simulated climate change (mimicking early snow melt) favors populations from lower elevations and snow addition restores the relative fitness advantage of high-elevation populations [@anderson_climate_2020]. Understanding what causes trait variation among populations within species, whether adaptive or not, will help biologists better predict how populations will respond to environmental change.

<!--- narrow problem: are phenotypic clines good evidence for local adaptation? --->
Phenotypic clines are commonly interpreted as evidence for local adaptation, but nonadaptive processes also generate similar patterns [@endler_geographic_1977]. A phenotypic cline is a correlation between a heritable character and the environment or a proxy for environmental gradients such as latitude or elevation [@huxley_clines_1938]. Clines may be caused by either abrupt or gradual environmental change. For example, heavy metal contamination of soil around mines is an abrupt change from the surrounding pasture soil. A similarly abrupt change in the heavy metal tolerance of *Anthoxanthum odoratum* occurs across the mine boundary, but other traits under weaker or correlated selection vary gradually over space [@antonovics_evolution_1970]. Other clines track gradual environmental change. Latitudinal clines in the size of *Drosophila subobscura* flies have evolved multiple times and most likely track gradual variation in temperature and/or phenology [@huey_rapid_2000]. But clines are not necessarily adaptive. When there is genetic isolation by distance [@wright_isolation_1943], it is possible for phenotypic differences to correlate with genetic differences throughout the species range [@vasemagi_adaptive_2006]. In practice, nonadaptive clines are difficult to demonstrate because it is hard to reject adaptive explanations. The proportion of cyanogenesis varies clinally in white clover (*Trifolium repens*) but field experiments have not observed spatially varying selection that could explain this cline [@wright_genetic_2021]. Hence, spatial variation in cyanogenesis may be nonadaptive or the design of the experiment may have missed selection occurring during germination and early development. When species expand their range, such as during biological invasions, nonadaptive clines can readily evolve because of multiple introductions and serial founder events [@colautti_contemporary_2015]. Even parallel clines, which are usually considered strong evidence of natural selection, can result from nonadaptive processes when there is epistasis and consistent spatial variation in the strength of genetic drift [@santangelo_modern_2018]. Evolutionary ecological experiments such as reciprocal transplants are needed to provide additional evidence to distinguish between adaptive and nonadaptive hypotheses for clines [@wadgymar_integrating_2017].

There are several experimental methods for measuring the agents and strength of selection to different environments under natural conditions [@wadgymar_identifying_2017], but here we focus on differential selection. One key prediction is that the direction of selection and position of the phenotypic optimum will differ between environments. If genetic variation in a trait *causes* fitness to vary, then genotypic selection analyses [@rausher_measurement_1992] should find that genotypes with high trait expression are favored in one environment but disfavored in another. For example, selection favors higher seed dormancy in southern environments of *Arabidopsis thaliana* and lower dormancy in northern environments, presumably to optimally time seedling establishment with the onset of favorable summer conditions [@postma_early_2016]. Populations can also simultaneously differ as a result of both adaptive and non-adaptive processes. When locally adapted populations differ in traits that do not themselves confer local adaptation, we expect population-level covariation between divergent traits and fitness but do not expect within-population variation to correlate with fitness. Hence, comparing the direction of among-genotype and among-population selection can help distinguish between phenotypic clines that are important for local adaptation and those that may have evolved nonadaptively.

In this study, we used a reciprocal transplant to test whether a cline in *Mimulus cardinalis* germination rate could be locally adaptive in terms of early survival between seedling establishment and flowering. Because seeds of *M. cardinalis* do not require cold or warm stratification to break dormancy, they will germinate whenever conditions are sufficiently warm and moist. Within this germination niche, there is genetic variation for germination time that covaries with latitude and other traits associated with a fast-slow growth continuum [@muir_grow_2017; @sheth_demographic_2018; @nelson_quantitative_2021]. In greenhouse conditions, seeds from the southern end of the range (San Diego County, CA) germinate in $6 \textendash 7$ days; seeds from the northern end of the range (Lane County, OR) germinate in $9 \textendash 10$ days. The growing season is typically longer in the southern end of the range because of reduced or nonexistent snowpack, but for the same reason southern populations may be more susceptible to late-season drought [@sheth_demographic_2018]. The adaptive hypothesis is that faster germination increases early survival in the southern end of the range and slower germination increases early survival at the northern end. We would predict either stabilizing genotypic selection for different optimal germination rates or opposing signs of directional genotypic selection. If the cline is maintained by nonadaptive forces then germination rate should be uncorrelated with genotypic fitness (even if it is correlated with population mean fitness) or the direction of selection should be the same in both parts of the range. A limitation of this study is that we cannot measure selection on seedling establishment since that occurred in the greenhouse (see [Materials and Methods]) or later survival and fecundity (these fitness components will be analyzed in a follow-up paper).

Germination and early survival are potentially important for local adaptation but often overlooked compared to later life stages. Failure to measure seedling establishment and early survival can bias estimates of selection on traits that only express at later life stages, such as flower size [@mojica_viability_2010]. In the herb *Arabidopsis thaliana*, quantitative differences in germination timing contributed to local adaptation and are maintained by a balance of selection on seedling establishment, survival, and fecundity, but the optimal timing varies in northern and southern environments [@postma_early_2016; @postma_among-year_2018]. A challenge of measuring genotypic selection on germination rate is that quantitative genetic theory and analysis are built around Gaussian distributions [@lynch_genetics_1998; but see @de_villemereuil_general_2016], but the distribution of times until germination are most likely not Gaussian. Although it is convenient to express germination as a rate (events per time), this and other "waiting time" traits are measured as the time until an event, the inverse of germination rate. Throughout this paper we distinguish between germination rate as the trait of interest which we estimate by measuring the time to germination. The time until germination (or another event) can be modeled as a waiting time, which is bounded at zero and usually right skewed. Waiting times are distributed exponentially if the probability of germination is constant through time, but this unlikely to be true. Rather, the probability of germination almost certainly increases through time once conditions are conducive. Hence, more complex waiting time distributions should occur in nature. A second challenge is that waiting time is usually modeled as a continuous trait, but is observed at discrete intervals (e.g. once per day). If the average time to germination is sufficiently long relative to the interval between observations, then a continuous-time approximation may be sufficient. However, when time to germination is several days, as with *M. cardinalis*, then accurately modeling the trait may need to account for the discrete observation process. Here we show that non-Gaussian distributions and measurement process can be incorporated into and improve analysis of time to germination. This approach should be useful, with some customization, to many other traits and biological systems. Hence, investigating spatially variable selection on germination time is important for expanding our knowledge of local adaptation and phenotypic clines for poorly studied life stages with challenging statistical properties.

@@

*Mimulus cardinalis* [$=$*Erythranthe cardinalis* @lowry_case_2019] is a perennial forb native to the Western US (California and Oregon). It is self-compatible, predominantly outcrossing, and hummingbird pollinated. We used five focal populations from throughout the geographic range of *M. cardinalis* (Table \ref{tab:focal_pops}; Fig. \ref{fig:range-map}). These five focal populations are a subset of those we used previously to identify a cline in germination and other traits [@muir_grow_2017]. Seeds were collected in the field from mature, undehisced fruits left in open coin envelopes for $2-4$ weeks to dry, then stored at room temperature. To minimize maternal effects, we grew a large number of field-collected seeds in the greenhouse and generated seed families for this experiment by hand-pollinating individuals using the breeding design described in the next section.

@@

tribble(
  ~`Source population`, ~Latitude, ~Longitude, ~`Elevation (mas)`,
  "Sweetwater River", "$32.900$", "$-116.585$", "$1180$",
  "West Fork Mojave River", "$34.284$", "$-117.378$", "$1120$",
  "North Fork Middle Tule River", "$36.201$", "$-118.759$", "$926$",
  "Little Jamison Creek", "$39.743$", "$-120.704$", "$1603$",
  "Rock Creek", "$43.374$", "$-122.957$", "$326$"
) |>
  knitr::kable(format = "markdown", caption = "\\label{tab:focal_pops} Source populations, including the name of the drainage where the seeds were collected, the latitude, longitude, and elevation in meters above sea level (mas).", align = "lccc", label = NA)
```

@@

To estimate selection among genotypes and populations, we first need to quantify genetic variance in germination rate and early survival. We estimated genetic variance and heritability of germination rate and winter survival using a quantitative genetic breeding design. For each population, we crossed $15$ parental individuals using a partial diallel design [@lynch_genetics_1998] with three dams per sire and three sires per dam for a total of $45$ full-sib families per population. The number of parents is low for quantitative genetic analyses, but we were restricted garden space and by the availability of wild-collected seeds in some source populations. We opted for fewer parental individuals and more populations so that we can better understand range-wide patterns of local adaptation. Individual plants of this hermaphroditic species were used as both sires (pollen parent) and dams (ovule parent). We did not make crosses between populations. One family from the Little Jamison Creek population did not produce enough seeds, resulting in $44$ usable families. In total, we used $5 \times 45 - 1 = 224$ families.

We estimated genetic variances and heritabilities for each of the five populations (Table \ref{tab:focal_pops}). The total phenotypic variance within a population $V_P$ can be partitioned into genetic ($V_G$), maternal ($V_M$), and environmental ($V_E$) components:

$$V_P = V_G + V_M + V_E$$
Genetic and environmental factors contribute to maternal affects [@lynch_genetics_1998]. Although we attempted to minimize maternal effects by growing a refresher generation in a common greenhouse environment and *M. cardinalis* seeds are tiny and not well provisioned, seed traits may be particularly sensitive to maternal effects. We did not have statistical power to estimate nonadditive genetic contributions, therefore we estimated broad-sense ($H^2$) heritability, the fraction of phenotypic variance contributed by all sources of (including dominance and epistasis) genetic variance:

@@

$$V_G = 4~\sigma_{\mathrm{parent}}$$
Using $\sigma_{\mathrm{parent}}$ rather than $\sigma_{\mathrm{sire}}$ is more powerful because it uses all of the data. We estimated $\sigma_{\mathrm{parent}}$ and other parameters using a Bayesian mixed effects model. We describe the general approach here and provide detail specific to germination rate and survival below. We fit the model in *Stan* version $`r stringr::str_c(fit[["metadata"]]()["stan_version_major"], fit[["metadata"]]()["stan_version_minor"], fit[["metadata"]]()["stan_version_patch"], sep = ".")`$ [@stan_development_team_stan_2022] using **cmdstanr** version $`r packageVersion("cmdstanr")`$ [@gabry_cmdstanr_2022]. *Stan* calculates the posterior distribution of the model using the Hamiltonian Monte Carlo algorithm, which is similar to the more widely used Markov Chain Monte Carlo, but is faster and more efficient at sampling for many applications [@monnahan_faster_2017]. We used weakly informative normally distributed priors for parameters that affect the trait mean (intercepts and coefficients) and half-Cauchy priors for variance parameters. Weakly informative priors are strongly recommended for complex mixed models [@mcelreath_statistical_2016]. We ran $`r length(fit$metadata()$id)`$ parallel chains for $`r fit[["metadata"]]()[["iter_warmup"]]`$ warmup iterations, $`r fit[["metadata"]]()[["iter_sampling"]]`$ sampling iterations, and a thinning interval of $`r fit[["metadata"]]()[["thin"]]`$. This configuration allowed parameters to converge, which we defined as the convergence diagnostic $\hat{R} < 1.05$ [@vehtari_rank-normalization_2021]. We inspected posterior predictive plots to assess model adequacy using the `pp_check` function from the *R* package **bayesplot** version $`r packageVersion("bayesplot")`$ [@gabry_visualization_2019]. We used the posterior median for point estimates and calculated uncertainty with the $95$\% quantile intervals from the posterior distribution. *Stan* and *R* code are available on GitHub ([https://github.com/cdmuir/mimulus-germination](https://github.com/cdmuir/mimulus-germination)) and will be archived on Zenodo upon publication. Data will be deposited on Dryad upon publication.

@@

We estimated germination rate under benign greenhouse conditions by measuring time to germination, the number of days to visible emergence from the soil. There were $48$ randomized blocks split evenly between two time cohorts (North cohort: August $3 \textendash 4$ and South cohort: August $22 \textendash 24$, $2015$) in the University of British Columbia greenhouse (Vancouver, BC, Canada). The North seed cohort went into the North transplant garden (gardens described below) and the South seed cohort went into the South transplant garden. Seed cohorts were staggered so that seedlings were transplanted at the same ontogenetic stage, as described below. We distinguish between cohorts and gardens because germination for each transplant garden was assayed in the greenhouse prior to transplanting. Emergence is a reliable proxy for germination since seeds were sown directly on top of the soil, resulting in little delay between germination and visible cotyledons. We sowed $3 \textendash 5$ seeds per plug on moist perlite-peat moss potting medium (Sunshine #1, Sungro Horticulture, Agawam, Massachusetts, USA) and recorded the first day on which a plug had at least one germinated seedling. We later thinned each plug to one seedling by selecting the central-most individual, so the first germinated seedling recorded is generally not the same individual transplanted into the field. Hence, individual correlations between germination and survival are inappropriate and we analyze relationships between breeding values or population averages. We censused germination daily for most of the experiment, but census dates were more spread out later in the experiment when only a few plants remained. We accounted for census interval in our statistical model (see below). Gentle misting kept soil moist for the duration of the experiment. Most plugs ($`r length(which(data[["Germinate"]] == "Y"))`$ of $`r nrow(data)`$, $`r round(100 * length(which(data[["Germinate"]] == "Y")) / nrow(data), 1)`$%) had at least one germinant by the end of the experiment (North cohort: August 27, 2015; South cohort: September 20, 2015); we treated plants that did not germinate as missing data. $`r nRemove`$ ($`r round(100 * nRemove / length(which(data[["Germinate"]] == "Y")), 2)`$%) plants emerged before day $5$. We removed data on these individuals because they were likely not *M. cardinalis*, but rather contamination from another species' seed in the potting medium. Species can be difficult to tell apart as the cotyledons first emerge, but in plugs where a single germinant emerged before day $5$ it was always a contaminant species. It therefore seems most likely that all germinants before day $5$ were contamination, but observations during thinning showed that contamination was rare and it unlikely to have any substantial impact on our estimate of germination timing.

## Estimating genetic variance in germination rate

In this section we describe the custom probability distribution we used to estimate quantitative genetic parameters of germination rate in this study. However, this approach could be extended to other ecologically important non-Gaussian traits censused at discrete temporal or spatial intervals. To modify the approach for other systems requires specifying an appropriate statistical distribution for the trait of interest and using integration to calculate the probability of an event occurring within a definite time or spatial interval. Probabilistic programming languages like *Stan* [@stan_development_team_stan_2022] have made it easier for biologists to specify statistical models most appropriate for their data.

The full germination-rate model included a fixed effect of population, fixed effect of cohort, a random effect of greenhouse block, a random maternal effect of dam, and a random effect of parent as either dam or sire to estimate genetic variance. Most quantitative traits are modeled with a Gaussian (i.e. normal) distribution, but this is inappropriate for germination which has a lower bound at $0$ and is censused at discrete time points. A Gaussian distribution might be an adequate approximation if the time to germination was longer and more spread out so that the effects of a zero lower bound and discrete sampling interval were inconsequential. However, given the rapid germination rate and small spread in timing for *M. cardinalis*, we used a different approach. Instead, we modeled genetic variation in a continuously distributed latent rate parameter that is measured as a discrete number of days to germination. Time to germination can be modeled similar to other waiting time or survival processes. There are several common waiting-time distributions with a lower bound at $0$, but we used the log-normal distribution because unlike some other common distributions (e.g. exponential), the mean and variance of log-normal distribution are separate parameters. Hence, we can model the variance components separate from differences in the mean germination rate. We discretized the distribution using the definite integral of the probability density to calculate a probability mass. Let the probability density with parameters $\vec{\theta}$ be $f(x|\vec{\theta})$. If a seed was censused to have germinated on day $t_i$, it could have germinated at any time between the previous census on day $t_{i-1}$. The associated probability mass $g(x|\theta)$ is:

$$g(x=t_i|\vec{\theta}) = \int_{t_{i-1}}^{t_i} f(x|\vec{\theta}) dx$$

Preliminary analyses revealed a germination threshold of four days since we did not observe any germination before the fifth day after sowing. The model did not fit as well without this threshold because models without a threshold predict a significant amount of germination between days $0$ and $5$ (results not shown). The variances and heritability are reported on the log-transformed scale.

@@

We compared the climate during the experiment (September $2015$ to May $2016$) to a standardized climate normal. Following @sheth_demographic_2018, we downloaded climate variables for $1961 \textendash 1990$ and $2015 \textendash 16$ derived from ClimateNA version $7.10$ [@wang_locally_2016]. We compared the seasonal temperature average (autumn: September-November; winter: December-February; spring: March-May) for the experiment with the $1961 \textendash 1990$ normal for each population.

@@

## Southern source populations germinate faster and have a higher probability of winter survival

@@

Fig. S1-S3 and Tables S1-S4 may be found online in the Supporting Information section.

@@

\begin{figure}[ht]
  \captionsetup{labelformat=empty}
  \caption{}
  \label{fig:range-map}
\end{figure}
